\section{Research Hypotheses}
\label{sec:research_objectives_research_hypothesis}

The main research hypotheses, corresponding to the research goals above, are:

\begin{itemize}
    \item \textbf{H1 (Unified Compositionality Hypothesis):}
    For any set of pretrained generative models, successful semantic 
    composition arises only when the composition operator 
    (energy-based product of experts, mixtures  , or density-based) is applied within a 
    representation that satisfies certain structural requirements.  
    When these requirements hold, the resulting composite distribution 
    corresponds to meaningful semantic combinations; when they fail, 
    each method exhibits predictable and characterisable failure modes.

    \item \textbf{H2 (Robust Guidance Hypothesis):}
    Property-guided compositional generation produces reliable and 
    semantically coherent samples only when the guidance model is 
    trained with mechanisms that ensure smoothness, calibrated 
    uncertainty, and robustness under distribution shift.  
    In particular, context-guided or uncertainty-aware training methods 
    are hypothesised to reduce hallucinated or implausible compositions 
    when steering composite densities toward continuous target values.

    \item \textbf{H3 (Intrinsic Geometric Compatibility Hypothesis).}
    Compositional correctness depends on the intrinsic geometric compatibility
    between the manifolds induced by pretrained diffusion models.We hypothesize that semantically valid
    composition is possible only when these induced geometries are compatible,
    meaning: (i) the manifolds are smoothly mappable or share a coherent joint
    structure; (ii) their local curvature is not in conflict. When these geometric conditions hold,
    compositional operators (energy-based product of experts, mixtures, density arithmetic) should yield 
    semantically plausible hybrid samples. When violated, we expect systematic failure
    modes associated with curvature mismatch, off-manifold sampling, or distorted
    anatomical structure.
\end{itemize}