% !TEX root = ../../proposal.tex

\section{Research Hypotheses}
\label{sec:research_objectives_research_hypothesis}

The main research hypotheses, corresponding to the research goals above, are:

\begin{itemize}
    \item \textbf{H1 (The Composability Gap Hypothesis):}
    Logical composition via additive score dynamics assumes a factorized joint distribution, which is only a valid proxy for semantic composition when the constituent concepts are conditionally independent.
    As the mutual information or relational dependency between concepts increases (e.g., spatial positioning or complex interactions), the logical composition fails to capture the emergent structural constraints of the true joint posterior.
    This results in a measurable divergence in latent trajectories, where logical composition enters regions of the manifold that are semantically unreachable through a single integrated prompt.
    \item \textbf{H2 (Pathology Independence Hypothesis):}
    To model spatial relation between concepts and ensure coherent attribute binding, the concepts being composed need to be factorized and independent for logical composition to represent semantic meaning.
    Furthermore, this assumption can be applied to real-world application such as medical imaging whereby pathologies are treated as concepts. If pathological disease conditions are treated as independent,
    then we can logically compose pre-trained diffusion models during inference to generate ecologically valid co-morbid conditions not seen during training.
    \item \textbf{H3 (Geometric Compatibility Hypothesis).}
    The composability gap demonstrates that logical composition lands off the natural data manifold thus producing low-quality and unrealistic outputs (hybridization).
    We hypothesize that natural data, such as medical images, occupy a riemannian manifold such that linear operations in the latent space fall off manifold.
    Hence, we expect logical composition using score-addition to fail to represent concepts semantically.
    We hypothesize that modeling the latent space as a riemannian manifold and traversing logical composition through geodesics keeps on the manifold such that
    composition retain semantic meaning. This closes the gap between logical and semantic composability.
\end{itemize}


H2: "If disease-related variation can be approximately factorised from shared anatomical and acquisition variation, then score composition in latent space should better approximate multi-pathology generation than composition in a fully entangled latent space
