% !TEX root = ../../eccv2016submission.tex

\section{Introduction}

Text-to-image diffusion models can produce coherent multi-object scenes from natural-language prompts, but they optimize for caption-level semantic plausibility rather than explicit logical constraints. This matters for compositional generation: prompting Stable Diffusion with ``$A$ and $B$'' and composing single-concept experts with an AND-like operator are not the same objective, even when both are intended to represent conjunction. In practice, the two paradigms can yield qualitatively different outputs, including semantically plausible co-occurrence scenes on one side and hybridized or dominant-mode outcomes on the other.

This paper studies that mismatch directly as a \emph{composability gap}. We compare semantic composition in SD~3.5~\cite{scaling_rectified_flow_transformers_for_high_resolution_image_synthesis} against logical score-space composition from composable diffusion~\cite{liu2022compositional} and SuperDiff AND~\cite{skreta2025the}. To isolate operator effects, all conditions start from the same noise sample per pair and seed. We then analyze both full denoising trajectories and terminal latents, so the gap is characterized not only by final visual differences but also by where and when trajectories separate.

We further test whether AND endpoints are reachable via standard text conditioning. For each SuperDiff-AND output, we project the image back into SD text-conditioning space using SD-IPC~\cite{ding2023sdipc}, a closed-form, training-free projection based on the pseudo-inverse of CLIP's text-projection matrix~\cite{ding2023sdipc}. This is a deliberately conservative choice: SD-IPC is the weakest plausible inverter, so failure here provides a lower-bound argument for structural inaccessibility. If even this minimal projection cannot close the cycle, then AND's target lies outside the region reachable by natural language conditioning.

Our experiments show a consistent pattern: semantic and logical composition diverge under shared-noise initialization, divergence is amplified in late denoising, and the SD-IPC cycle breaks for AND outputs while closing for monolithic text generation. These findings establish the composability gap as a structural property, not a prompting artefact, and open a research programme: understanding its representational cause (why do separately trained experts produce off-manifold compositions?) and designing geometric corrections that enforce manifold consistency during inference.
