% !TEX root = ../../../proposal.tex

\section{Generative Models for Chest Radiography}
\label{sec:medical_imaging_background}

\subsection{Pathology-Conditioned Synthesis}

Chest radiography is a high-volume, low-cost modality central to the diagnosis of
cardiopulmonary disease. The VinBigData dataset~\citep{nguyen2022vindr} provides
15{,}000 posteroanterior (PA) radiographs annotated by professional radiologists
across 14 pathological findings, following a pronounced long-tail distribution in
which normal cases comprise 70.7\% of all images. This scarcity of annotated
pathological cases --- and the near absence of jointly annotated co-morbid cases ---
creates a persistent data augmentation challenge. Generative models have been
explored as a solution: by synthesising pathology-conditioned images, they can
supplement real training data in low-resource settings. Early work used conditional
GANs~\citep{gan_medical_image_augmentation_2018, medical_image_synthesis_gans_2018}
to produce disease-specific images, with demonstrated downstream benefits for
classification in small-data regimes. More recently, latent diffusion
models~\citep{rombach2022high} have produced higher-quality and more diverse
radiograph synthesis, with conditioning provided by text or class embeddings. These
monolithic conditioning approaches generate a single disease at a time; they do not
address the problem of synthesising rare co-morbid presentations in which two or more
pathologies must coexist coherently in the same image.

\subsection{Domain-Specific Representations}

General-purpose vision-language models such as CLIP are trained on natural image--text
pairs and are not calibrated to the visual statistics of radiographs. Domain-specific
pre-training addresses this gap. CheSS (Chest X-ray Self-supervised Contrastive
Learning)~\citep{NEURIPS2022_123d3e81} is an encoder pre-trained on large-scale chest
radiograph collections using a contrastive objective, producing embeddings in which
radiologically similar images are closer than radiologically dissimilar ones.
A domain-specific encoder provides better-calibrated similarity measurements for
conditioning and evaluation in chest X-ray generation tasks than general-purpose
vision encoders. In Phase~3, CheSS embeddings are used to condition the per-disease
diffusion experts and to evaluate the radiological fidelity of composed co-morbid
samples.
