% !TEX root = ../../proposal.tex

\begin{abstract}
Compositional generative modeling aims to create complex unseen combinations by composing multiple pretrained models. 
Instead of relying on a single large-scale system, individual expert models can be combined at inference time to enable out-of-distribution generation. 
This approach avoids the need to retrain a new model for every concept pairing. 
This capability is critical in low-resource settings where specific data combinations are rare. 
In clinical imaging, for example, hospitals often have many scans for single diseases but very few examples of rare co-morbidities. 
A compositional system can bridge this gap by synthesizing plausible rare cases from these separate components to support data augmentation.

However, current compositional methods often fail. 
These systems struggle with tasks like attribute binding, negation, and combining closely related concepts. 
A primary reason for this failure is the mathematical operator used to combine conditions, which we define as logical composition. 
This mathematical approach does not always match the semantically meaningful output produced by a single prompt for a complex scene. 
This creates a disconnect between the mathematical combination and the intended scene. 
The result limits image quality and makes synthetic data unreliable for sensitive clinical use.

This study investigates this disconnect and develops a method to reduce it. 
The research follows three steps. First, it characterizes the failure of logical composition through a new taxonomy of concept pairs and trajectory analysis. 
Second, it tests whether learning more factorized representations, using a partitioned latent space, can align the mathematical combinations with the target scene. 
Finally, it demonstrates the utility of this framework in medical imaging. The study evaluates whether rare, co-morbid chest X-rays can be synthesized from individual common and separate disease models. 
This work aims to make out-of-distribution synthesis more reliable and useful for practical downstream tasks.
\end{abstract}
