\chapter{Research Plan}
\section{Two-Year Research Trajectory}
\subsection{Months 1--4: Foundations and Reproducible Infrastructure}
During the first semester I will consolidate the proposal, co-design a detailed measurement plan with my supervisors, and complete the Wits HPC onboarding required to access the \texttt{bigbatch}, \texttt{stampede}, and \texttt{biggpu} partitions. This period will be used to finalise ethics clearances, refresh statistical methodology, and containerise the current baseline diffusion pipeline so that experiments can be repeated across partitions. Weekly meetings with Prof.~Klein and Dr.~Jarvis will be used to prioritise milestones and schedule early consultations with the Wits Digital Health group to align expectations about clinically useful outputs.
\subsection{Months 5--10: Data Curation and Baseline Modelling}
With governance and infrastructure in place, effort shifts to curating the radiography corpora and engineering baseline diffusion models. Data harmonisation and augmentation jobs will be submitted to \texttt{bigbatch} to take advantage of its throughput while exploratory model training and ablation studies will use \texttt{stampede}. Frequent checkpoints (every six weeks) with hospital collaborators will verify that preprocessing choices preserve clinically salient cues. The outcome is a calibrated baseline model, associated evaluation scripts, and a reproducible report describing strengths, weaknesses, and open questions.
\subsection{Months 11--16: Compositional Diffusion Development}
The second project year begins with the introduction of compositional conditioning strategies, curriculum schedules, and disentangled latent variables. I will prototype components locally before scaling combined experiments on \texttt{biggpu}. Hyper-parameter searches will be automated through workflow managers so that runs can be resumed after maintenance windows. Interim findings will be disseminated in an internal lab talk and a targeted workshop submission to obtain early peer feedback.
\subsection{Months 17--20: Evaluation, Interpretation, and Specialist Validation}
Once the best-performing configurations have stabilised, I will quantify robustness, fairness, and generalisability across demographic subgroups. Quantitative experiments will mix \texttt{stampede} for batched inference with \texttt{biggpu} for attention visualisation and counterfactual sampling. I will run structured evaluation sessions with radiologists and methodologists to interpret examples, triage failure modes, and co-develop acceptance criteria for clinical deployment. Their feedback will directly feed into model refinements and the drafting of an evaluation chapter.
\subsection{Months 21--24: Synthesis and Dissemination}
The final semester is dedicated to consolidating insights, addressing outstanding experiments, and packaging artefacts. A reproducibility bundle containing code, trained checkpoints, and experiment manifests will be archived on the \texttt{stampede} partition and mirrored to institutional storage. I will prepare journal and conference manuscripts, integrate feedback from supervisory and clinical partners, and finalise the thesis chapters. Dry-run defences with the supervisory team and external mentors will ensure that the narrative is coherent and that claims are defensible.

\section{Experimentation, Interpretation, and Specialist Engagement}
Every experiment will be launched via version-controlled manifests so that changes to datasets, hyper-parameters, or software dependencies are recorded. Results will be synchronised into a central experiment tracker that tags runs with the executing partition, allowing hardware-specific anomalies to be detected quickly. Quantitative summaries will be complemented with qualitative panels of generated radiographs, uncertainty estimates, and saliency overlays to make interpretations accessible to clinicians. Monthly reading groups with the Wits Biomedical Engineering unit and quarterly advisory boards with partner hospitals will provide recurring opportunities to verify findings, correct biases, and collect domain-specific advice. Insights from these engagements will be translated into actionable changes that are tested in the subsequent experimental cycle, ensuring rapid validation loops.

\section{Risk Management and Contingency Planning}
Potential risks include data access delays, queue saturation on preferred partitions, and shifts in clinical partner availability. To mitigate them I will: (i) maintain fallback datasets and synthetic augmentation pipelines that can be executed locally; (ii) diversify job submissions across \texttt{bigbatch}, \texttt{stampede}, and \texttt{biggpu} depending on their availability; (iii) document experiments comprehensively so that progress can continue asynchronously when collaborators are unavailable; and (iv) schedule periodic buffer weeks for refactoring and documentation. These contingencies ensure that the project stays on track even under adverse conditions.
\input{chapters/research_plan/dataset_collection}
% \input{chapters/research_plan/timeline}